---
title: "Associations Between Variables"
format:
  html:
    toc: true
    embed-resources: true
    code-tools: true
editor: 
  markdown: 
    wrap: 72
---

## Overview

This lesson covers four types of variable association, each with its own
visualisation approach, statistical test, and regression framing:

| Predictor type | Outcome type | Test family | Plot |
|----|----|----|----|
| Categorical (2 groups) | Continuous | t-test | Boxplot |
| Categorical (3+ groups) | Continuous | ANOVA | Boxplot / histogram |
| Categorical | Categorical | Chi-squared | Bar / mosaic |
| Continuous | Continuous | Correlation / regression | Scatter |

::: callout-note
**All of these can be unified under the linear model framework using
`lm()`**.
:::

------------------------------------------------------------------------

## Setup

```{r load-packages}
#| message: false
# install.packages("pak")  # uncomment and run once if pak is not installed
pacman::p_load(
  tidyverse,
  ggpubr,        # stat_compare_means() and stat_cor() for annotating plots
  infer,         # tidy statistical tests on dataframes
  vcd,           # chi-squared residual shading in mosaic plots
  broom,         # tidy() to convert test output into dataframes
  moderndive,    # readable regression tables: get_regression_*()
  marginaleffects, # modern effects plots, replacing sjPlot
  gapminder,      # gapminder dataset
  pak        # to load ggmosaic
)

# ggmosaic is not on CRAN — install from GitHub once, then load normally
pak::pak("haleyjeppson/ggmosaic")
library(ggmosaic)
```

```{r load-data}
#| message: false
# Load from local file if available — otherwise fall back to the web
if (file.exists("data/random_qs.csv")) {
  dat <- read_csv("data/random_qs.csv")
} else {
  dat <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQul_g9GtsSj-30sQ2hOE_rixxy8gH6AdhawcT5AoRJodJdGZnBgqtwgDAmfab0CJ1AjmFCrKmVw5MA/pub?gid=1799592268&single=true&output=csv")

  dat <- dat |>
    set_names(c("timestamp", "height", "month", "date", "visit1", ">14",
                "guess1", "visit2", "<96", "guess2", "data_sci")) |>
    mutate(
      visit        = coalesce(visit1, visit2),
      visit        = str_detect(visit, "Yes"),
      guess        = coalesce(guess1, guess2),
      data_sci     = str_detect(data_sci, "Yes"),
      timestamp    = lubridate::mdy_hms(timestamp),
      date         = parse_number(date),
      odd          = date %% 2 == 1
    ) |>
    unite("birthday", c(month, date), sep = " ") |>
    mutate(birthday_24 = str_c(birthday, " 2024") |> lubridate::mdy()) |>
    select(timestamp, height, birthday, visit, guess, data_sci, odd, birthday_24)

  write_csv(dat, "data/random_qs.csv")
}
```

```{r load-resume}
#| message: false
resume <- read_csv("https://tinyurl.com/iss-racial-disc-web")
```

```{r load-gapminder}
gapminder <- gapminder::gapminder
```

### Checking and cleaning the height data

Before any analysis, it is good practice to check for anomalies in the
data. The `arrange()` call below sorts by height ascending so that any
suspicious values appear at the top.

```{r check-height}
dat |> arrange(height)
```

::: callout-note
## Question

What looks wrong with the height data? What are two plausible
explanations for the anomalous value(s)?
:::

There are two reasonable strategies for dealing with the problem:

```{r height-fix-options}
#| eval: false
# Option A: simply filter out anything implausibly small
dat |>
  filter(height > 7) |>
  arrange(height)

# Option B: assume values under 100 are in metres and convert,
# then replace anything still implausible with NA
dat |>
  mutate(height = ifelse(height < 100, height * 100, height)) |>
  mutate(height = na_if(height, height > 200))
```

We use Option B to preserve as many observations as possible, saving the
result as a new working dataframe `df`.

```{r create-df}
df <- dat |>
  mutate(height = ifelse(height < 100, height * 100, height)) |>
  mutate(height = na_if(height, 700))
```

------------------------------------------------------------------------

## Part 1 — Continuous \~ categorical: birthday parity and Africa guess

The first question is whether people born on an odd-numbered date give
different guesses for the number of countries in Africa compared with
those born on an even-numbered date. This is a classic two-group
comparison: one continuous outcome variable (`guess`) and one binary
categorical predictor (`odd`).

### Visualisation

A boxplot is the natural first step — it shows the median, spread, and
any outliers for each group side by side.

```{r boxplot-guess-odd}
ggplot(df) +
  aes(x = "", y = guess, fill = odd) +
  geom_boxplot() +
  scale_fill_hue() +
  labs(x = "Birthday date",
       y = "Guess: number of countries in Africa",
       title = "Africa guess by odd/even birthday") +
  theme_minimal() +
  ggpubr::stat_compare_means(method = "t.test")  # use label.x and label.y to control label placement
```

::: callout-note
## Questions

-   Do the plots suggest a difference between the two groups?
-   `stat_compare_means()` adds the test result directly to the plot.
    What does the p-value tell you here?
:::

### t-test

The t-test asks whether the difference in means between two groups is
larger than would be expected by chance. The base-R `t.test()` works on
vectors, so we first split the dataframe into two groups and extract the
relevant column.

```{r t-test-split}
df_odd  <- df |> filter(odd)
df_even <- df |> filter(!odd)
```

```{r t-test-base}
t.test(df_even$guess, df_odd$guess)
```

The output gives us the mean for each group and a 95% confidence
interval for the difference. If that interval crosses zero, we cannot
rule out that there is no real difference. We can pipe the result into
`broom::tidy()` for a cleaner single-row table.

```{r t-test-tidy}
t.test(df_even$guess, df_odd$guess) |>
  broom::tidy()
```

The columns mean:

-   `estimate` — difference in means between the two groups
-   `estimate1` / `estimate2` — mean of group 1 and group 2 respectively
-   `statistic` — the t-score
-   `parameter` — degrees of freedom
-   `conf.low` / `conf.high` — bounds of the 95% confidence interval

The `infer` package lets us run the same test directly on the dataframe
without splitting it first, using a formula interface that matches
`lm()`.

```{r t-test-infer}
df |>
  infer::t_test(formula = guess ~ odd)
```

### Regression framing

Any t-test result can be reproduced exactly using a linear regression
model. This matters because `lm()` generalises to many more situations
than `t.test()` does.

```{r lm-guess-odd}
cat_model <- lm(guess ~ odd, data = df)
summary(cat_model)
```

```{r lm-guess-odd-moderndive}
get_regression_summaries(cat_model)
get_regression_table(cat_model)
```

The regression table tells us:

-   `(Intercept)` estimate — predicted value for the base category (even
    birthdays), equivalent to the group mean
-   `oddTRUE` estimate — predicted *change* when moving from even to odd
-   `statistic` — t-value (identical to the t-test above)
-   `lower_ci` / `upper_ci` — 95% confidence interval around each
    estimate

```{r lm-guess-odd-points}
get_regression_points(cat_model)
```

`guess_hat` shows the model's predicted value for each observation (the
group mean), and `residual` shows how far the actual response was from
that prediction.

------------------------------------------------------------------------

## Part 2 — Continuous \~ multi-categorical: continent and life expectancy

When the categorical predictor has more than two levels, we move from a
t-test to an ANOVA (Analysis of Variance). The logic is the same — we
are asking whether group means differ more than chance would predict —
but now we are comparing several groups simultaneously.

```{r gapminder-2007}
gapminder2007 <- gapminder |>
  filter(year == 2007) |>
  select(country, lifeExp, continent, gdpPercap)

glimpse(gapminder2007)
```

### Visualisation

```{r histogram-lifeexp-continent}
ggplot(gapminder2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5, color = "white") +
  labs(x = "Life expectancy",
       y = "Number of countries",
       title = "Distribution of life expectancies by continent (2007)") +
  facet_wrap(~continent, nrow = 2)
```

```{r boxplot-lifeexp-continent}
ggplot(gapminder2007, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  labs(x = "Continent", y = "Life expectancy",
       title = "Life expectancy by continent (2007)") +
  ggpubr::stat_compare_means(method = "anova") +
  theme_minimal()
```

::: callout-note
## Question

The ANOVA p-value tells you whether *any* group differs from the others
— but not *which* groups differ. What follow-up test would you use to
find out?
:::

```{r summary-lifeexp-continent}
lifeExp_by_continent <- gapminder2007 |>
  group_by(continent) |>
  summarise(median = median(lifeExp),
            mean   = mean(lifeExp))
lifeExp_by_continent
```

### Regression framing

```{r lm-lifeexp-continent}
lifeExp_model <- lm(lifeExp ~ continent, data = gapminder2007)
summary(lifeExp_model)
get_regression_summaries(lifeExp_model)
get_regression_table(lifeExp_model)
get_regression_points(lifeExp_model, ID = "country")
```

::: callout-note
## Questions

-   Which continent is the base (reference) category in the regression
    output, and how can you tell?
-   What does the estimate for `continentEurope` represent?
:::

### Effects plot

Rather than reading the table of estimates, an effects plot shows the
model's predicted values with confidence intervals — much easier to
interpret visually.

```{r effects-lifeexp-continent}
marginaleffects::plot_predictions(lifeExp_model, condition = "continent") +
  labs(x = "Continent", y = "Predicted life expectancy",
       title = "Model predictions: life expectancy by continent (2007)") +
  theme_minimal()
```

------------------------------------------------------------------------

## Part 3 — Categorical \~ categorical

When both variables are categorical, we use a contingency table and a
chi-squared test. The question is whether the *joint* distribution of
two variables differs from what we would expect if they were
independent.

### Birthday parity and Japan visit

A quick sanity check: is there any association between birthday parity
and whether a respondent has visited Japan? There should not be — this
is a useful null result to practise with.

```{r contingency-odd-visit}
# Frequency counts for each variable individually
table(df$odd)
table(df$visit)

# Contingency table: rows = odd/even, columns = visited Japan
table(df$odd, df$visit)
```

```{r chisq-odd-visit}
# Chi-squared test via summary() on the table
summary(table(df$odd, df$visit))
```

::: callout-note
## Question

The output includes the message "Chi-squared approximation may be
incorrect." What causes this warning, and what alternative test does the
script use instead?
:::

```{r fisher-odd-visit}
# Fisher's exact test — more reliable when cell counts are small
fisher.test(df$odd, df$visit)
```

```{r chisq-infer-odd-visit}
# Tidy alternative with infer
df |>
  infer::chisq_test(formula = odd ~ visit)
```

### Visualising the joint distribution

```{r bar-odd-visit}
ggplot(df) +
  aes(x = visit, fill = odd) +
  geom_bar() +
  scale_fill_viridis_d(option = "cividis") +
  coord_flip() +
  theme_minimal()
```

```{r mosaic-odd-visit}
ggplot(df) +
  geom_mosaic(aes(x = product(visit, odd), fill = odd)) +
  labs(x = "Visited Japan", y = "Odd birthday",
       title = "Birthday parity vs Japan visit") +
  theme_mosaic() +
  theme(legend.position = "none")
```

::: callout-note
## `vcd::mosaic()` and Pearson residuals

`geom_mosaic()` fits naturally into a ggplot workflow, but it does not
show which cells drive a significant chi-squared result. For that,
`vcd::mosaic(shade = TRUE)` is uniquely useful: it colours tiles by
their **Pearson residual**, which measures how far each observed cell
count is from what we would expect if the two variables were completely
independent.

$$r = \frac{\text{observed} - \text{expected}}{\sqrt{\text{expected}}}$$

-   **Blue tiles** (positive residual) — the cell occurs *more* often
    than independence predicts
-   **Red tiles** (negative residual) — the cell occurs *less* often
    than independence predicts
-   **Pale/unshaded tiles** — the cell is close to what independence
    would expect

The colour intensity reflects the size of the residual. Cells with \|r\|
\> 2 are noteworthy; \|r\| \> 4 is striking. This makes `vcd::mosaic()`
particularly useful *after* a significant chi-squared result, to
identify which specific combinations of categories are driving it.

```{r mosaic-odd-visit-vcd}
vcd::mosaic(~odd + visit,
            direction = c("v", "h"),
            data = df,
            shade = TRUE)
```

The pale tiles here confirm what the chi-squared test already told us:
birthday parity and Japan visit are not meaningfully associated.
:::

------------------------------------------------------------------------

### Real-world example: racial discrimination in the labour market

> Bertrand & Mullainathan (2004) sent fictitious résumés to employers in
> response to newspaper ads. The résumés were identical in all respects
> except the applicant's name — some were stereotypically associated
> with Black candidates, others with White candidates. The outcome
> variable was whether the employer called back. This design allows any
> difference in callback rates to be attributed to the name alone, i.e.
> to racial discrimination.

```{r glimpse-resume}
glimpse(resume)
```

Note that `call` is stored as a numeric 0/1 variable. We will need to
convert it to a factor or logical before using it with `infer`.

```{r contingency-resume}
# Raw counts
table(resume$race, resume$call)

# Percentages, rounded to one decimal place
round(prop.table(table(resume$race, resume$call)) * 100, 1)

# Chi-squared test
summary(table(resume$race, resume$call))
```

```{r chisq-resume-infer}
#| error: true
# This errors because infer expects a categorical response variable
resume |>
  infer::chisq_test(call ~ race)
```

::: callout-note
## Question

Read the error message carefully. What does it tell you about how
`infer` expects the response variable to be stored? How do we fix it?
:::

```{r chisq-resume-infer-fixed}
resume |>
  mutate(call = as.factor(call)) |>
  infer::chisq_test(call ~ race)
```

### Visualising the résumé data

```{r bar-resume-numeric}
# First attempt — call is numeric, so ggplot treats it as a continuous scale
resume |>
  ggplot(aes(x = race, fill = call)) +
  geom_bar() +
  scale_fill_viridis_d(option = "cividis") +
  coord_flip() +
  theme_minimal()
```

::: callout-note
## Question

The plot above does not look right. Run `class(resume$race)` and
`class(resume$call)`. What is the problem, and why does converting
`call` to logical fix it?
:::

```{r bar-resume-logical}
resume |>
  mutate(call = as.logical(call)) |>
  ggplot(aes(x = race, fill = call)) +
  geom_bar() +
  scale_fill_viridis_d(option = "cividis") +
  coord_flip() +
  theme_minimal()
```

```{r mosaic-resume}
ggplot(resume |> mutate(call = as.logical(call))) +
  geom_mosaic(aes(x = product(call, race), fill = call)) +
  labs(x = "Race", y = "Called back",
       title = "Callback rate by race") +
  theme_mosaic()
```

Now let's add residual shading to see which cells are driving the
significant chi-squared result.

```{r mosaic-resume-vcd}
# shade = TRUE colours tiles by Pearson residual
# Blue = more callbacks than independence would predict
# Red  = fewer callbacks than independence would predict
vcd::mosaic(~race + call,
            direction = c("v", "h"),
            data = resume,
            shade = TRUE)
```

The blue tile for White names × called back and the red tile for Black
names × called back show precisely where the disparity lies: White
applicants were called back *more* than independence would predict,
while Black applicants were called back *less*. The residuals make the
direction of discrimination visible at a glance, in a way that the
chi-squared p-value alone cannot.

### Regression framing

Because `call` is coded as 0/1, we can treat it as a numeric outcome and
fit `lm()` — the results are equivalent to the chi-squared test.

```{r lm-resume}
cat_cat_model <- lm(call ~ race, data = resume)
summary(cat_cat_model)
get_regression_summaries(cat_cat_model)
get_regression_table(cat_cat_model)
```

The regression table tells us:

-   `(Intercept)` estimate — the mean callback rate for the base
    category (Black names)
-   `racewhite` estimate — the *additional* callback probability for
    White names
-   The F-statistic from `get_regression_summaries()` is approximately
    equal to the chi-squared statistic from the table above

------------------------------------------------------------------------

::: callout-tip
## DIY challenge — sex and racial discrimination

Using `dplyr`, create a new dataframe `res_2` that combines `race` and
`sex` into a single `applicant` column (e.g. "black_female",
"white_male").

Steps: 1. Take the `resume` dataframe 2. `group_by(race, sex)` 3.
`mutate()` a new column called `applicant` using `paste()` to combine
the strings 4. `ungroup()` 5. `select()` only `applicant` and `call`

Then use a chi-squared test and a mosaic plot to ask: is there evidence
of racial discrimination for both sexes?
:::

------------------------------------------------------------------------

## Part 4 — Continuous \~ continuous: correlation and regression

When both variables are continuous, we describe their relationship with
a correlation coefficient and model it with a regression line.

### Height and Africa guess

Does a person's height predict their guess for the number of African
countries? Almost certainly not — but it is a useful dataset to practise
with since we already know it well.

```{r scatter-height-guess}
ggplot(df) +
  aes(x = height, y = guess) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor(method = "spearman") +
  theme_minimal()
```

```{r cor-test-height-guess}
cor.test(df$height, df$guess, method = "pearson") |>
  broom::tidy()
```

The output columns mean:

-   `estimate` — the correlation coefficient (r), ranging from −1 to 1
-   `statistic` — the t-score used to test whether r differs from zero
-   `parameter` — degrees of freedom
-   `conf.low` / `conf.high` — 95% confidence interval around r

If the confidence interval crosses zero, we cannot rule out that there
is no real relationship.

::: callout-note
## Question

The confidence interval for the height–guess correlation crosses zero.
What does this mean for our conclusion about the relationship? Note also
that correlation tests assume the variables are approximately normally
distributed. How would you check that assumption before running the
test?
:::

```{r lm-height-guess}
cont_model <- lm(formula = guess ~ height, data = df)
summary(cont_model)
get_regression_summaries(cont_model)
get_regression_table(cont_model)
get_regression_points(cont_model)
```

The regression line can be drawn manually using the intercept and slope
estimates from `get_regression_table()`. Extending the x-axis to zero
makes the intercept visible.

```{r scatter-height-guess-extended}
ggplot(df) +
  aes(x = height, y = guess) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlim(0, 200) +
  ggpubr::stat_cor(method = "pearson") +
  theme_minimal()
```

::: callout-tip
## DIY

Add `geom_abline()` to the plot above, using the intercept and slope
values from `get_regression_table(cont_model)`. Does the line match
`geom_smooth()`?
:::

------------------------------------------------------------------------

### Real-world example: GDP per capita and life expectancy

```{r cor-gdp-lifeexp}
# Pearson correlation (assumes linearity and normality)
cor(gapminder$gdpPercap, gapminder$lifeExp)

# Kendall's tau — a rank-based alternative for non-normal data
cor(gapminder$gdpPercap, gapminder$lifeExp, method = "kendall")

# Full test with confidence intervals
cor.test(gapminder$gdpPercap, gapminder$lifeExp, method = "pearson") |>
  broom::tidy()
```

```{r scatter-gdp-lifeexp}
gapminder |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor(label.y = 130) +
  theme_minimal()
```

::: callout-note
## Question

The regression line does not fit the data well at the lower end of GDP.
Before reading on, look at the scatter plot and think about what
transformation might improve the fit.
:::

Checking the distributions of both variables reveals the problem:
`gdpPercap` is heavily right-skewed, violating the normality assumption
of the Pearson correlation.

```{r histograms-gdp-lifeexp}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram() +
  theme_minimal()

ggplot(gapminder, aes(x = gdpPercap)) +
  geom_histogram() +
  theme_minimal()
```

A log transformation pulls the long right tail in and makes the
distribution much more symmetric.

```{r histogram-log-gdp}
ggplot(gapminder, aes(x = log2(gdpPercap))) +
  geom_histogram() +
  theme_minimal()
```

```{r scatter-log-gdp-lifeexp}
gapminder |>
  ggplot(aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  theme_minimal()
```

::: callout-note
## Question

The R value increases from 0.58 to 0.81 after the log transformation.
What does this tell you about the relationship between GDP and life
expectancy — is it linear or does it follow a different curve?
:::

### Two regression models compared

```{r lm-gdp-lifeexp-raw}
gdp_life_model01 <- lm(lifeExp ~ gdpPercap, data = gapminder)
summary(gdp_life_model01)
get_regression_summaries(gdp_life_model01)
get_regression_table(gdp_life_model01)
```

```{r gapminder-log}
gapminder_log <- gapminder |>
  mutate(log_gdp = log(gdpPercap))
```

```{r lm-gdp-lifeexp-log}
gdp_life_model02 <- lm(lifeExp ~ log_gdp, data = gapminder_log)
summary(gdp_life_model02)
get_regression_summaries(gdp_life_model02)
get_regression_table(gdp_life_model02)
get_regression_points(gdp_life_model02)
```

------------------------------------------------------------------------

## Package summaries

The sections below provide a consolidated reference for each package
used in this lesson. Use them as a cheat sheet when working on your own
analyses.

### `moderndive` — readable regression tables

`moderndive` wraps `lm()` output into three clean functions. First, fit
your models:

```{r package-summary-models}
# Rebuild all models for the summary section
cat_model       <- lm(guess ~ odd, data = df)
multi_cat_model <- lm(lifeExp ~ continent, data = gapminder2007)
cat_cat_model   <- lm(call ~ race, data = resume)
cont_model      <- lm(guess ~ height, data = df)
gdp_log_model   <- lm(lifeExp ~ log_gdp, data = gapminder_log)
```

`get_regression_summaries()` — overall model fit:

```{r moderndive-summaries}
get_regression_summaries(cat_model)
get_regression_summaries(multi_cat_model)
get_regression_summaries(cat_cat_model)
get_regression_summaries(cont_model)
get_regression_summaries(gdp_log_model)
```

-   `r_squared` — correlation coefficient (closer to 1 = stronger
    relationship)
-   `adj_r_squared` — adjusted for the number of predictors
-   `mse` / `rmse` / `sigma` — measures of model error (lower is better)
-   `statistic` — F-statistic
-   `p_value` — is the overall model significant?
-   `nobs` — number of observations

`get_regression_table()` — coefficient estimates:

```{r moderndive-tables}
get_regression_table(cat_model)
get_regression_table(multi_cat_model)
get_regression_table(cat_cat_model)
get_regression_table(cont_model)
get_regression_table(gdp_log_model)
```

-   First row (`Intercept`) — predicted value for the base category, or
    when all continuous predictors equal zero
-   Subsequent rows — predicted *change* for each one-unit increase
    (continuous) or shift from base to that category (categorical)
-   `lower_ci` / `upper_ci` — 95% confidence interval around each
    estimate

`get_regression_points()` — observed vs fitted values:

```{r moderndive-points}
get_regression_points(cat_model)
get_regression_points(multi_cat_model, ID = "country")
get_regression_points(cat_cat_model)
get_regression_points(cont_model)
get_regression_points(gdp_log_model)
```

-   `_hat` columns — the model's predicted value for each observation
-   `residual` — the difference between the predicted and actual value

------------------------------------------------------------------------

### `infer` — tidy statistical tests

`infer` lets you run statistical tests directly on a dataframe using a
formula interface, avoiding the need to extract vectors manually.

```{r package-summary-infer}
# t-test
df |>
  infer::t_test(formula = guess ~ odd)

# chi-squared test (response variable must be a factor)
resume |>
  mutate(call = factor(call)) |>
  infer::chisq_test(call ~ race)
```

------------------------------------------------------------------------

### `broom` — tidy output from base-R tests

`broom::tidy()` converts the messy list output of base-R test functions
into a clean single-row dataframe, making it easy to compare results or
include them in a report.

```{r package-summary-broom}
# t-test
t.test(df_even$guess, df_odd$guess) |>
  broom::tidy()

# correlation test
cor.test(df$height, df$guess, method = "pearson") |>
  broom::tidy()

# linear model coefficients
lm(guess ~ odd, data = df) |>
  broom::tidy()

# ANOVA
aov(formula = gdpPercap ~ continent, data = gapminder2007) |>
  broom::tidy()
```

------------------------------------------------------------------------

### `ggmosaic` and `vcd` — mosaic plots

Two packages produce mosaic plots, and they complement each other well.

**`ggmosaic::geom_mosaic()`** is a proper ggplot geom, so it accepts the
usual `theme_*()`, `scale_*()`, and `labs()` layers and sits
consistently alongside the rest of a ggplot-based analysis. The key
syntax difference from other geoms is `aes(x = product(var1, var2))`,
which tells ggmosaic how to divide the plot into tiles.

```{r package-summary-ggmosaic}
ggplot(resume |> mutate(call = as.logical(call))) +
  geom_mosaic(aes(x = product(call, race), fill = call)) +
  labs(x = "Race", y = "Called back") +
  theme_mosaic()
```

**`vcd::mosaic(shade = TRUE)`** adds something `geom_mosaic()` cannot
easily replicate: automatic **Pearson residual shading**. After a
significant chi-squared test, this tells you not just *that* there is an
association, but *which specific cells* are responsible for it.

The Pearson residual for each cell is:

$$r = \frac{\text{observed} - \text{expected}}{\sqrt{\text{expected}}}$$

where "expected" is the count you would predict if the two variables
were completely independent. The colour legend in the plot shows the
residual thresholds: by convention, \|r\| \> 2 is noteworthy and \|r\|
\> 4 is striking.

-   **Blue** = more observations than independence predicts
    (over-represented)
-   **Red** = fewer observations than independence predicts
    (under-represented)
-   **Pale** = close to what independence would expect

```{r package-summary-vcd}
vcd::mosaic(~race + call,
            direction = c("v", "h"),
            data = resume,
            shade = TRUE)
```

A practical workflow: use `geom_mosaic()` for exploratory visualisation
and reporting (it blends with your other ggplots), then switch to
`vcd::mosaic(shade = TRUE)` after a significant chi-squared result to
diagnose which cells are driving it.

------------------------------------------------------------------------

### `marginaleffects` — effects plots

`marginaleffects` is the modern replacement for `sjPlot::plot_model()`.
Rather than jumping straight to a plot, it first produces a tidy
dataframe of predicted values that you can inspect, then pass to ggplot
if you want a custom figure. `plot_predictions()` is a convenience
wrapper that does both steps at once.

```{r package-summary-marginaleffects}
# Categorical predictor: predicted group means with confidence intervals
plot_predictions(cat_model, condition = "odd") +
  theme_minimal()

# Multi-categorical predictor
plot_predictions(multi_cat_model, condition = "continent") +
  theme_minimal()

# Categorical ~ categorical: predicted callback rates by race
plot_predictions(cat_cat_model, condition = "race") +
  theme_minimal()

# Continuous predictor: predicted regression line with confidence band
plot_predictions(cont_model, condition = "height") +
  theme_minimal()

# Log-transformed continuous predictor
plot_predictions(gdp_log_model, condition = "log_gdp") +
  theme_minimal()
```

If you want the underlying predicted values as a dataframe rather than a
plot, use `predictions()` instead:

```{r marginaleffects-predictions-df}
#| eval: false
marginaleffects::predictions(multi_cat_model,
                             newdata = datagrid(continent = unique)) |>
  as_tibble()
```

------------------------------------------------------------------------

### `ggpubr` — annotating plots with test results

`ggpubr::stat_compare_means()` adds a statistical test result directly
to a ggplot as a text annotation.

```{r package-summary-ggpubr-ttest}
# Two-group comparison: specify method = "t.test" for a parametric test
ggplot(df) +
  aes(x = "", y = guess, fill = odd) +
  geom_boxplot() +
  scale_fill_hue() +
  labs(x = "Birthday date",
       y = "Guess: countries in Africa",
       title = "Africa guess by odd/even birthday") +
  theme_minimal() +
  ggpubr::stat_compare_means(method = "t.test")

# Omitting method uses the non-parametric Wilcoxon test by default — safer
# when the normality assumption is uncertain
ggplot(df) +
  aes(x = "", y = guess, fill = odd) +
  geom_boxplot() +
  scale_fill_hue() +
  labs(x = "Birthday date",
       y = "Guess: countries in Africa",
       title = "Africa guess by odd/even birthday") +
  theme_minimal() +
  ggpubr::stat_compare_means()
```

```{r package-summary-ggpubr-anova}
# Multi-group comparison with ANOVA
# With multiple x categories, label.x refers to the category index (1, 2, 3...)
# so label.x = 2 places the label above the second continent
ggplot(gapminder2007, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  labs(x = "Continent", y = "Life expectancy",
       title = "Life expectancy by continent") +
  ggpubr::stat_compare_means(method = "anova", label.x = 4, label.y = 45) +
  theme_minimal()

# Default non-parametric (Kruskal-Wallis) version
ggplot(gapminder2007, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  labs(x = "Continent", y = "Life expectancy",
       title = "Life expectancy by continent") +
  ggpubr::stat_compare_means(label.x = 4, label.y = 45) +
  theme_minimal()
```

::: callout-note
## Parametric vs non-parametric tests

The parametric versions (t-test, ANOVA) assume the data are
approximately normally distributed. When that assumption is uncertain —
especially with small samples — the non-parametric alternatives
(Wilcoxon, Kruskal-Wallis) are safer because they make no distributional
assumptions. As a rule of thumb, use non-parametric tests unless you
have a good reason to believe normality holds.
:::
