---
title: "Schedule"
sidebar: false
toc: false
page-layout: full
freeze: false
---

```{r packages-data, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
withr::with_dir(here::here(), {
  schedule_nested <- targets::tar_read(schedule_page_data)
  targets::tar_load(schedule_ical_file)
})
base_url <- yaml::read_yaml(here::here("_quarto.yml"))$website$`site-url`
ical_url_absolute <- paste0(base_url, "/", schedule_ical_file)
```

::: {.schedule-page}

Here's your roadmap for the semester!

- [**Content**](/content/index.qmd) ({{< fa book-open-reader >}}): This page contains the session outline, readings, and R code examples for each week. Start here.

- [**Videos**](/example/index.qmd) ({{< fa laptop-code >}}): Supporting videos from YouTube to help with key concepts and coding in R. These are especially useful when working through assignments.

- [**Assignments**](/assignment/index.qmd) ({{< fa pen-ruler >}}): Weekly exercises using the `r4np` interactive package. Assignments include a short weekly reflection on what you have learned. **Assignments are due by 11:59 PM on the Sunday after each lesson.** See the schedule table below to see exactly when each one is due.

- [**Projects**](/project/index.qmd) ({{< fa diagram-project >}}): Problem sets and the data project. Problem sets (PS02–PS06) are available roughly once every two weeks and are **optional extra credit** — they build deeper statistical skills using real datasets. The data project is the **main course assessment** and replaces a final exam.

:::::: {.row .d-flex .justify-content-center}
::::: {.col-12 .col-sm-10 .col-md-6}
:::: {.card .bg-danger .text-white}
::: {.card-body}

**tl;dr**: Follow this general process each week:

- Work through the content page ({{< fa book-open-reader >}})
- Complete the assignment ({{< fa pen-ruler >}}) using the videos ({{< fa laptop-code >}}) as a reference
- Work on problem sets ({{< fa diagram-project >}}) when available for extra credit

:::
::::
:::::
::::::

---

::: {.callout-tip}
## Subscribe!

You can subscribe to this calendar URL in Outlook, Google Calendar, or Apple Calendar:

::: {.grid}
::: {.g-col-12 .g-col-md-2}
<p class="download-btn"><a class="btn btn-danger btn-sm" href="/`r schedule_ical_file`" role="button">{{< fa calendar-alt >}} Download</a></p>
:::
::: {.g-col-12 .g-col-md-10}
<p><input class="form-control" type="text" value="`r ical_url_absolute`" onclick="this.select()" readonly></p>
:::
:::
:::

```{r build-table, include=FALSE}
show_table <- function(group_id) {
  # Add a heading
  cat(as.character(paste("\n\n###", schedule_nested$group[[group_id]], "\n\n")))
  
  # Make the table
  tbl <- schedule_nested$data[[group_id]] |> 
    kbl(escape = FALSE, align = "rlcccc", table.attr = 'class="schedule-table"') |> 
    kable_styling() |> 
    column_spec(1, width = "20%", extra_css = "padding-right: 20px;") |> 
    column_spec(2, width = "48%") |> 
    column_spec(3:6, width = "8%")
  
  cat(tbl)
}
```

```{r show-table, echo=FALSE, results="asis"}
walk(seq(1, nrow(schedule_nested)), ~show_table(.x))
```

:::
