---
title: "Problem Set 03: Data Wrangling"
author: "WRITE YOUR NAME HERE"
date: today
format:
  html:
    toc: true
    toc-float: true
    embed-resources: true
    df-print: kable
    theme: lumen
    code-download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

In this problem set we will practise the key data manipulation tasks for describing, summarising, and working with data. We will specifically review the following functions from the `dplyr` package:

- `select()` — extract columns
- `filter()` — subset rows by condition
- `mutate()` — create new variables
- `summarize()` — compute summary statistics
- `arrange()` — sort rows
- `group_by()` — group data before summarising
- `|>` — the pipe operator

---

## Setup

Run this code block first.

```{r libraries}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(ggplot2, dplyr)
```

## The data

```{r load-data, eval=FALSE}
data(txhousing)
glimpse(txhousing)
```

These data are about housing in Texas. Each row is monthly data for a given city in a given year.

**Question 1.** Open the data viewer (`View(txhousing)` in the console, or click the name in the Environment pane). What is the last city listed in the dataset (row 8602)?

**Answer:**

**Question 2.** Type `?txhousing` in the console. What information does the `listings` variable contain?

**Answer:**

---

## Data wrangling review

### select()

Extract specific columns. The `-` sign means "all columns *except* this one":

```{r select-example}
txhousing <- txhousing |> select(-date)
```

### filter()

Keep only rows matching a condition:

```{r filter-example}
austin_12 <- txhousing |> filter(year == 2012, city == "Austin")
```

Use `|` for "or":

```{r filter-or}
aust_ab <- txhousing |> filter(city == "Austin" | city == "Abilene")
```

### mutate()

Add new columns:

```{r mutate-example}
txhousing <- txhousing |>
  mutate(vol_100k = volume / 100000)
```

### summarize()

Reduce a column to a summary statistic:

```{r summarize-example}
austin_12 |> summarize(x_bar_sales = mean(sales),
                        sd_sales    = sd(sales),
                        min_vol     = min(volume),
                        max_vol     = max(volume),
                        mdn_list    = median(listings),
                        iqr_list    = IQR(listings),
                        sample_size = n())
```

### arrange()

Sort rows in descending order:

```{r arrange-example, eval=FALSE}
austin_12 |>
  arrange(desc(volume))
```

### group_by()

Calculate summary statistics per group:

```{r groupby-example}
aust_ab |>
  group_by(city) |>
  summarize(x_bar_sales = mean(sales))
```

---

## Independent practice

### Basic syntax

**Question 3.** Write a code chunk to remove the `inventory` variable from `txhousing`. Save the result back as `txhousing`.

```{r q3}
# Your code here

```

**Question 4.** Make a dataset called `dallas_sub` that includes data only from Dallas in 2012 and 2013.

```{r q4}
# Your code here

```

**Question 5.** Add a column to `dallas_sub` called `prct_sold` that calculates the percentage of listings that were sold (`sales / listings * 100`). Save the result back as `dallas_sub`.

```{r q5}
# Your code here

```

**Question 6.** Calculate the **average** percentage of listings sold in Dallas **in each month of the year** using `dallas_sub`. Save the result as `dallas_summary`.

```{r q6}
# Your code here

```

**Question 7.** Arrange `dallas_summary` in descending order by average percentage sold. Which month had the greatest percentage of houses sold?

```{r q7}
# Your code here

```

**Answer:**

---

### More advanced wrangling

**Question 8.** Run the code below, study it, and explain in your own words what it calculates.

```{r q8, eval=FALSE}
txhousing |>
  filter(year == 2012 | year == 2013, city == "Dallas") |>
  mutate(prct_sold = sales / listings * 100) |>
  group_by(month) |>
  summarize(mean_prct_sold = mean(prct_sold)) |>
  arrange(desc(mean_prct_sold))
```

**Answer:**

**Question 9.** In January 2015, what city had the fewest houses listed for sale?

```{r q9}
# Your code here

```

**Answer:**

**Question 10.** In 2012, in which month were the most houses sold in Texas?

```{r q10}
# Your code here

```

**Answer:**

**Question 11.** Generate a single table showing the total number of houses sold in **Austin** in 2000 and 2001 combined, and the total number sold in **Dallas** in 2000 and 2001 combined.

*Hint: `sum()` adds values.*

```{r q11}
# Your code here

```

---

*When you are done, render this document to HTML and submit on Moodle. Name your file `PS03_yourname.html`.*
