---
title: "Problem Set 05: Linear Regression — One Categorical Predictor"
author: "WRITE YOUR NAME HERE"
date: today
format:
  html:
    toc: true
    toc-float: true
    embed-resources: true
    df-print: kable
    theme: lumen
    code-download: true

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 16/2, fig.height = 9/2)
```

## Background

In this problem set we model data on hate crimes recorded across US states in the 10 days after the 2016 presidential election, as measured by the Southern Poverty Law Center (SPLC).

For context, see the FiveThirtyEight article: [Higher Rates of Hate Crimes Are Tied to Income Inequality](https://fivethirtyeight.com/features/higher-rates-of-hate-crimes-are-tied-to-income-inequality/) (January 2017).

We will use regression models with a **single categorical** explanatory variable.

---

## Setup

```{r libraries}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(ggplot2, dplyr, moderndive, readr)
```

## The data

```{r load-data}
hate_crimes <- read_csv("data/hate_crimes.csv")
```

```{r glimpse, eval=FALSE}
glimpse(hate_crimes)
```

Each row is a US state. The outcome variable is `hate_crimes` (hate crimes per 100,000 individuals). We examine three categorical explanatory variables:

- `trump_support` — level of Trump support: `low`, `medium`, or `high`
- `unemployment` — level of unemployment: `low` or `high`
- `median_house_inc` — median household income: `low` or `high`

---

## Part 1: Hate crimes and Trump support

### Question 1A

Create a visualisation to explore the relationship between `hate_crimes` and `trump_support`. Include axis labels and a title.

Note: due to alphabetical ordering, the baseline (reference) group is `high`. We will keep this ordering as-is throughout.

```{r q1a}
# Your code here

```

### Question 1B

Comment on the relationship between these two variables. Is this what you would have expected?

**Answer:**

### Question 2

Fit a regression model for `hate_crimes ~ trump_support` and generate the regression table.

```{r q2}
# Your code here

```

### Question 3A

What does the intercept represent in this regression table?

**Answer:**

### Question 3B

What does the model estimate as the number of hate crimes per 100,000 people in states with **low** Trump support?

**Answer:**

### Question 3C

Does the model estimate that hate crimes are more frequent in states with **low** or **high** Trump support?

**Answer:**

### Question 3D

How much greater were hate crimes in **medium** Trump-support states compared to **high** Trump-support states?

**Answer:**

### Question 3E

What are the three possible fitted values $\hat{y}$ for this model? Use `get_regression_points()` to explore if unsure.

```{r q3e, eval=FALSE}
# Your code here

```

**Answer:**

### Question 3F

The code below calculates the group means of `hate_crimes` for each level of `trump_support`. How do these compare to the three possible fitted values from Question 3E?

```{r q3f}
hate_crimes %>%
  group_by(trump_support) %>%
  summarize(mean_hate_crimes = mean(hate_crimes, na.rm = TRUE))
```

**Answer:**

---

The regression equation for this model is:

$$\hat{y} = 0.191 + 0.269 \times \mathbf{1}_{\text{low}}(x) + 0.031 \times \mathbf{1}_{\text{med}}(x)$$

### Question 4

Solve the regression equation for a state where `trump_support` is **high**.

**Answer:**

### Question 5

Which 5 states had the highest rate of hate crimes? What was the level of Trump support in these 5 states? Do these results surprise you?

```{r q5}
# Your code here

```

**Answer:**

---

## Part 2: Hate crimes and unemployment

### Question 6

Create a visualisation to explore the relationship between `hate_crimes` and `unemployment`. Include axis labels and a title.

```{r q6}
# Your code here

```

### Question 7A

Fit a regression model for `hate_crimes ~ unemployment` and generate the regression table.

```{r q7a}
# Your code here

```

### Question 7B

Write out the regression equation for this model.

**Answer:**

### Question 7C

What does the intercept represent in this regression table?

**Answer:**

### Question 7D

What does the model estimate as the number of hate crimes per 100,000 people in states with **low** unemployment?

**Answer:**

### Question 7E

What are the two possible fitted values $\hat{y}$ for this model? Why are there only two this time instead of three?

**Answer:**

### Question 8

Use `get_regression_points()` to generate a table of fitted values and residuals for the `hate_crimes ~ unemployment` model. Examine the first row: how are the residuals calculated?

```{r q8, eval=FALSE}
# Your code here

```

**Answer:**

---

## Part 3: Hate crimes and household income

### Question 9A

Fit a regression model for `hate_crimes ~ median_house_inc` and generate the regression table.

```{r q9a}
# Your code here

```

### Question 9B

Were there more hate crimes in states with **high** or **low** median household incomes? How large was the difference?

**Answer:**

### Question 10

Run `get_regression_points()` for the `hate_crimes ~ median_house_inc` model and examine the data for Maine (row 2). Did the model **overpredict** or **underpredict** the hate crime level for Maine?

```{r q10, eval=FALSE}
# Your code here

```

**Answer:**

---

*When you are done, render this document to HTML and submit on Moodle. Name your file `PS05_yourname.html`.*
