---
title: "Problem Set 06: Multiple Regression"
author: "WRITE YOUR NAME HERE"
date: today
format:
  html:
    toc: true
    toc-float: true
    embed-resources: true
    df-print: kable
    theme: lumen
    code-download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 16/2, fig.height = 9/2)
```

## Background

We will again use the hate crimes data from Problem Set 05. This week we fit regression models with **both** a numerical and a categorical explanatory variable.

The outcome variable this week is `income` — the median household income in each US state. The explanatory variables are:

- `hs` — percentage of adults aged 25+ with a high school degree (numerical)
- `urbanization` — level of urbanization in the state: `low` or `high` (categorical)

---

## Setup

```{r libraries}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(ggplot2, dplyr, moderndive, readr)
```

## The data

```{r load-data}
hate_crimes <- read_csv("data/hate_crimes.csv")
```

```{r glimpse, eval=FALSE}
glimpse(hate_crimes)
```

---

## Income, education, and urbanization

### Question 1

Create a visualisation comparing median household `income` at **low** and **high** levels of urbanization (do not include `hs` in this plot). Include axis labels and a title.

```{r q1}
# Your code here

```

### Question 2A

Create a scatterplot showing:

- `income` on the y axis
- `hs` on the x axis
- Points coloured by `urbanization`
- **Parallel** regression lines using `geom_parallel_slopes()` from the `moderndive` package

```{r q2a}
# Your code here

```

### Question 2B

Do you think the relationship between `hs` and `income` is strong or weak? Linear or non-linear?

**Answer:**

### Question 2C

Which regression line (high or low urbanization) appears to have the larger intercept?

**Answer:**

### Question 3A

Create the same scatterplot again, but this time use `geom_smooth(method = "lm")` to draw separate, non-parallel regression lines for each urbanization group.

```{r q3a}
# Your code here

```

### Question 3B

How do the slopes in Question 3A compare to the parallel slopes in Question 2A? Are they fairly similar or quite different?

**Answer:**

### Question 3C

Based on the two plots, do you think a **parallel slopes** model or an **interaction model** would be more appropriate here?

**Answer:**

---

### Question 4A

Fit both of the following models:

1. A parallel slopes model: `income ~ hs + urbanization`
2. An interaction model: `income ~ hs * urbanization`

Save the output of each `lm()` call.

```{r q4a}
# Your code here

```

### Question 4B

Use `get_regression_summaries()` to find the unadjusted R² for each model. Report both values.

```{r q4b}
# Your code here

```

**Answer:**

### Question 4C

Compare the **adjusted** R² for each model. Which model do you prefer? Does this agree with your visual assessment in Question 3C?

**Answer:**

---

*For Questions 5–9, base your answers on your preferred model from Question 4.*

### Question 5A

Generate the regression table using `get_regression_table()`. Is the intercept the same for high and low urbanization states? Is the slope the same?

```{r q5a}
# Your code here

```

**Answer:**

### Question 5B

What is the slope and intercept for states with **high** urbanization?

**Answer:**

### Question 5C

What is the slope and intercept for states with **low** urbanization?

**Answer:**

### Question 5D

Is median household income greater in states with lower or higher urbanization? By how much?

**Answer:**

### Question 5E

For every 1 percentage point increase in high-school educated adults, what is the associated increase in median household income?

**Answer:**

### Question 6

Predict the median household income for a state with **high** urbanization where 85% of adults have a high school degree.

**Answer:**

### Question 7

Predict the median household income for a state with **low** urbanization where 85% of adults have a high school degree.

**Answer:**

### Question 8

Predict the median household income for a state with **low** urbanization where 30% of adults have a high school degree.

**Answer:**

### Question 9

Use `get_regression_points()` to look at the data for Maine (row 2). What was Maine's observed `income`? What did your model predict? What is the residual? Did the model over- or underestimate?

```{r q9}
# Your code here

```

**Answer:**

---

## Independent analysis: vole habitat

Wildlife biologists want to understand habitat preferences of a declining vole species. They counted voles at 56 random study sites and recorded:

- `voles` — vole count at each site
- `veg` — percent cover of vegetation
- `soil` — whether the site is `moist` or `dry`

```{r load-voles}
vole_trapping <- read_csv("http://bit.ly/2IgDF0E")
```

Fit an appropriate regression model with `voles` as the outcome variable and `veg` and `soil` as explanatory variables. Use the results to answer Questions 10–12.

```{r vole-model}
# Your code here

```

### Question 10

Would protecting a site with high vegetation cover be a more effective way to preserve the vole population than a site with low vegetation cover? Why?

**Answer:**

### Question 11

Dry sites are cheaper to purchase and maintain. If a conservation organisation buys dry sites, roughly what minimum percent cover of vegetation do they need to maintain to support a population of about 30 voles?

**Answer:**

### Question 12

The Nature Conservancy is considering a site with **moist** soil and **40% vegetation cover**. Using the regression equation, what vole population might this site support?

**Answer:**

---

*When you are done, render this document to HTML and submit on Moodle. Name your file `PS06_yourname.html`.*
