---
title: "Monte Carlo Simulation & Descriptive Statistics"
format: html
editor: source
---

## Overview

This lesson covers three connected topics:

1. **Sampling and Monte Carlo simulation** — using `sample()`, `replicate()`, and custom functions to simulate dice rolls and explore the Central Limit Theorem
2. **Descriptive statistics** — exploring a real class height dataset alongside official Japanese survey data
3. **Classic Monte Carlo examples** — estimating π and simulating the Birthday Paradox

---

## Setup

```{r load-packages}
#| message: false
pacman::p_load(tidyverse, gridExtra,skimr, modelsummary)
```

```{r load-data}
#| message: false
# Japanese height survey data
jpn_ht <- read_csv("https://tinyurl.com/iss-jpn-height-web")

# Class questionnaire data — live Google Sheet export
dat <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQul_g9GtsSj-30sQ2hOE_rixxy8gH6AdhawcT5AoRJodJdGZnBgqtwgDAmfab0CJ1AjmFCrKmVw5MA/pub?gid=1799592268&single=true&output=csv")
```

```{r rename-columns}
# Rename to something manageable
names(dat) <- c("completed", "height", "month", "date", "visit1",
                ">14", "n_africa1", "visit2", "<96", "n_africa2", "iss_datasci")
```

```{r wrangle-data}
#| message: false
# Merge split columns, derive birthday parity, convert logicals, tidy selection
dat <- dat |>
  mutate(visit      = coalesce(visit1, visit2),
         n_africa   = coalesce(n_africa1, n_africa2),
         birthday   = ifelse(is.na(visit1), "even", "odd"),
         iss_datasci = iss_datasci == "Yes") |>
  select(completed, iss_datasci, height, month, date, birthday, visit, n_africa)

# Convert month names to numbers
dat <- dat |>
  mutate(month = match(month, month.name))
```

::: {.callout-tip}
## Challenge — explore the dataframe
- What are the three ways to call `View()`?
- What is another function (not `View()`) that gives you a quick look at a dataframe?
:::

```{r save-data}
#| eval: false
# Save both datasets locally — only needs to run once
# After this, you can read from data/ instead of the web
write_csv(jpn_ht, "data/jpn_ht.csv")
write_csv(dat,    "data/africa_guess.csv")
```

---

## Part 1 — Sampling and replicating

Monte Carlo simulation (also called bootstrapping) uses repeated random sampling
to explore probability. It is the conceptual foundation of many machine learning methods.

### 1.1 `sample()` basics

```{r sample-names}
# Change these to match the actual class members!
names <- c("member_1", "member_2", "member_3", "member_4")

# Pick two names without replacement
sample(names, size = 2, replace = FALSE)

# Pick five with replacement (same name can come up more than once)
sample(names, size = 5, replace = TRUE)
```

::: {.callout-note}
## Question
Results will differ each time because sampling is random. How do we make random
results reproducible? What does `set.seed()` actually do?
:::

```{r sample-set-seed}
set.seed(111)
sample(names, size = 2, replace = FALSE)
sample(names, size = 5, replace = TRUE)
```

### 1.2 Simulating dice

```{r dice-single}
# Rolling one die
sample(1:6, 1)

# Rolling two dice
sample(1:6, size = 2, replace = TRUE)
```

::: {.callout-note}
## Question
Why do we need `replace = TRUE` when rolling two dice, but not when picking
two names from a list?
:::

```{r dice-sum}
# Save the roll and sum it
roll <- sample(1:6, size = 2, replace = TRUE)
roll
sum(roll)

# Same in one line
sum(sample(1:6, size = 2, replace = TRUE))
```

### 1.3 Functions and `replicate()`

```{r two-dice-function}
# A function that rolls two dice and returns their sum
two_dice <- function() {
  roll <- sample(1:6, size = 2, replace = TRUE)
  return(sum(roll))
}
two_dice()
```

```{r replicate-intro}
#| eval: false
# replicate() runs a function n times and returns the results as a vector
# Here we use getwd() as a harmless example
replicate(n = 3, getwd())
```

::: {.callout-note}
## Question
`replicate()` is shown here with `getwd()` as a safe demonstration.
What does `getwd()` return, and why is it a useful choice for illustrating `replicate()`?
:::

```{r replicate-dice}
# Roll two dice four times
replicate(n = 4, two_dice())

# Save as a vector and summarise
scores <- replicate(n = 4, two_dice())
scores

mean(scores)
median(scores)
range(scores)   # lowest and highest values
table(scores)   # frequency table
sd(scores)
hist(scores)    # quick base-R histogram
```

::: {.callout-tip}
## DIY 1
Change the `n` value below and observe how the histogram changes.
What trend do you notice as `n` gets larger?
:::

```{r diy-replicate}
#| eval: false
scores <- replicate(n = ____, two_dice())
hist(scores)
```

---

## Part 2 — Plotting simulations

### 2.1 Why ggplot needs a dataframe

```{r ggplot-error}
#| eval: false
# This errors because ggplot expects a dataframe, not a vector
ggplot(data = scores, aes(x = scores)) +
  geom_histogram()
```

::: {.callout-note}
## Question
What is the exact error message produced above, and what does it tell you
about how ggplot works differently from base-R `hist()`?
:::

```{r scores-as-df}
# Wrap scores in a tibble alongside names to use with ggplot
scores <- replicate(n = 4, two_dice())
df <- tibble(name  = names,
             score = scores)

# Histogram
ggplot(df, aes(x = score)) +
  geom_histogram(binwidth = 1, colour = "white") +
  scale_x_continuous(breaks = 0:12, limits = c(1, 12))

# Bar chart — one bar per person
ggplot(df, aes(x = name, y = score)) +
  geom_col() +
  scale_y_continuous(breaks = 0:12, limits = c(0, 12))
```

### 2.2 A simulation function

```{r test-2-dice-function}
# Wraps replicate() and two_dice() into a single function that returns a df
test_2_dice <- function(n) {
  tibble(roll  = 1:n,
         score = replicate(n, two_dice()))
}

t_5    <- test_2_dice(5)
t_50   <- test_2_dice(50)
t_5000 <- test_2_dice(5000)

mean(t_5$score)
mean(t_50$score)
mean(t_5000$score)

median(t_5$score)
median(t_50$score)
median(t_5000$score)
```

::: {.callout-note}
## Question
As `n` increases, what happens to the mean and median? What statistical principle
does this illustrate?
:::

```{r pipe-into-ggplot}
# Because test_2_dice() returns a df, you can pipe straight into ggplot
test_2_dice(20) |>
  ggplot() +
  aes(x = score) +
  geom_histogram(binwidth = 1, colour = "white")
```

::: {.callout-note}
## Question
The original script uses `%>%` here rather than `|>`. What is the difference
between the two pipe operators? Does it matter here?
:::

### 2.3 A plotting function

```{r plot-2-dice-function}
plot_2_dice <- function(n) {
  test <- tibble(throw = 1:n,
                 score = replicate(n, two_dice()))
  ggplot(test) +
    aes(x = score) +
    geom_histogram(binwidth = 1L, fill = "#0c4c8a", colour = "white") +
    scale_x_continuous(breaks = 0:12, limits = c(0, 14)) +
    ggtitle(paste("Histogram after", n, "rolls")) +
    theme_minimal()
}
```

::: {.callout-tip}
## DIY
Call `plot_2_dice()` with different values of `n`. What trend do you notice
as `n` increases? What shape does the distribution approach?
:::

```{r plot-2-dice-diy}
#| eval: false
plot_2_dice(____)
```

### 2.4 Comparing with the normal distribution

```{r norm-2-dice-function}
norm_2_dice <- function(n) {
  test <- tibble(throw = 1:n,
                 score = replicate(n, two_dice()))

  # Histogram
  p1 <- ggplot(test) +
    aes(x = score) +
    geom_histogram(binwidth = 1L, fill = "#0c4c8a", colour = "white") +
    scale_x_continuous(breaks = 0:12, limits = c(0, 14)) +
    ggtitle(paste("Histogram after", n, "rolls")) +
    theme_minimal()

  # Density plot overlaid with a theoretical normal distribution
  p2 <- ggplot(test) +
    aes(x = score) +
    geom_density(fill = "#0c4c8a", alpha = 0.5,
                 adjust = logb(n, 50)) +  # bandwidth scales with sample size
    scale_x_continuous(breaks = 0:12, limits = c(0, 14)) +
    ggtitle("Density plot vs Normal distribution") +
    theme_minimal() +
    stat_function(colour = "red",
                  fun = dnorm,
                  args = list(mean = mean(test$score), sd = sd(test$score)))

  grid.arrange(p1, p2, nrow = 1)
}
```

::: {.callout-note}
## Questions
- What does `adjust = logb(n, 50)` do to the density bandwidth? Why scale it with `n`?
- What does `stat_function()` add to the plot? What arguments does `dnorm` need?
:::

```{r norm-2-dice-demo}
norm_2_dice(10)
```

```{r norm-2-dice-map}
# Run norm_2_dice() for several values of n and stack the results
rolls <- as.list(c(5, 50, 500, 5000, 50000))
plots <- map(rolls, ~ norm_2_dice(.x))
do.call(grid.arrange, c(plots, list(ncol = 1)))
```

::: {.callout-note}
## Questions
- What does `map()` do here? Why does `rolls` need to be a list?
- What does `do.call(grid.arrange, ...)` do? Why can't you just call `grid.arrange(plots)`?
- What pattern do you see across the five plots as `n` grows?
:::

---

## Part 3 — Exploring the class height data

### 3.1 Subsetting and summary statistics

```{r subset-data}
df_all   <- dat
df_class <- dat |> filter(iss_datasci)  # only this class's responses
```

```{r summary-dollar}
# Extracting a column as a vector using $
df_class$height
mean(df_class$height)
median(df_class$height)
max(df_class$height)
min(df_class$height)
var(df_class$height)
sd(df_class$height)
IQR(df_class$height)
```

```{r summary-attach}
# attach() lets you refer to columns by name without the df$ prefix
# Useful for exploratory work on a single df — but always detach() afterwards!
attach(df_class)

mean(height)
median(height)
max(height)
min(height)
var(height)
sd(height)
IQR(height)

detach(df_class)
```

::: {.callout-note}
## Question
What are the risks of using `attach()`? Why is it important to call `detach()` when finished?
:::

```{r height-summary-df}
attach(df_class)

class_height_summary <- tibble(
  mean   = mean(height),
  median = median(height),
  max    = max(height),
  min    = min(height),
  var    = var(height),
  sd     = sd(height),
  IQR    = IQR(height)
)
class_height_summary

detach(df_class)
```
You can also use a base R function to get a `summary` of your data:

```{r summary-base}
summary(df_class)
```

::: {.callout-note}
## Going further — richer summaries
Base R's `summary()` is a good starting point, but specialist packages give you
more detail and better formatting. Try these two alternatives and compare what
each adds.
:::

```{r summary-skimr}
#| eval: false
# install if needed: install.packages("skimr")
skimr::skim(df_class)
```

```{r summary-modelsummary}
#| eval: false
# install if needed: install.packages("modelsummary")
modelsummary::datasummary_skim(df_class)
```

### 3.2 Frequency tables

```{r frequency-tables}
attach(df_class)

# Raw frequency count — note the alphabetical default ordering
table(month)

# Better: make month a factor with meaningful level order
table(factor(month, levels = month.name))

# Proportions — note: prop.table() needs a table object, not a raw factor
round(prop.table(table(month)), 2)

detach(df_class)
```

::: {.callout-note}
## Questions
- Why does `prop.table(month)` produce an error but `prop.table(table(month))` works?
- The comment says functions "work from the centre out, like in Excel" — what does this mean?
:::

::: {.callout-tip}
## Challenge
Repeat the summary statistics and frequency table exploration above using `df_all`
instead of `df_class`. Do you notice anything odd?
:::

---

### 3.3 Visualising height — class data

```{r boxplot-class}
ggplot(df_class) +
  aes(x = "", y = height, fill = birthday) +
  geom_boxplot() +
  scale_fill_hue() +
  labs(x = "Birthday", y = "Height in cm",
       title = paste("Boxplot of height by odd/even birthday —",
                     nrow(df_class), "responses")) +
  theme_minimal()
```

::: {.callout-note}
## Question
The title shows the sample size. Is `n` large enough to draw a reliable conclusion
about whether birthday parity is related to height?
:::

```{r density-class-combined}
ggplot(df_class) +
  aes(x = height) +
  geom_density(adjust = 0.5, fill = "#0c4c8a") +
  labs(x = "Height in cm", y = "Probability Density",
       title = paste("Density plot of class height, n =", nrow(df_class))) +
  theme_minimal()
```

::: {.callout-note}
## Question
The density plot looks bimodal (two peaks). What might explain this shape
in a class of university students?
:::

```{r density-class-by-birthday}
ggplot(df_class) +
  aes(x = height, fill = birthday) +
  geom_density(adjust = 0.5, alpha = 0.5) +
  geom_rug(aes(colour = birthday)) +
  labs(x = "Height in cm", y = "Probability Density",
       title = paste("Density plot by birthday parity, n =", nrow(df_class))) +
  theme_minimal()
```

::: {.callout-note}
## Question
What does `geom_rug()` add? When is it useful compared to just using the density curve?
:::

### 3.4 Visualising height — full dataset and outlier removal

```{r density-all-combined}
ggplot(df_all) +
  aes(x = height) +
  geom_density(adjust = 1, fill = "#0c4c8a") +
  labs(x = "Height in cm", y = "Probability Density",
       title = paste("Density plot of all responses, n =", nrow(df_all))) +
  theme_minimal()
```

```{r density-all-by-birthday}
ggplot(df_all) +
  aes(x = height, fill = birthday) +
  geom_density(adjust = 1, alpha = 0.5) +
  geom_rug(aes(colour = birthday)) +
  labs(x = "Height in cm", y = "Probability Density",
       title = paste("Height by birthday parity — all responses, n =", nrow(df_all))) +
  theme_minimal()
```

```{r boxplot-all-with-outlier}
ggplot(df_all) +
  aes(x = "", y = height, fill = birthday) +
  geom_boxplot() +
  scale_fill_hue() +
  labs(x = "Birthday", y = "Height in cm",
       title = paste("Boxplot — all responses, n =", nrow(df_all))) +
  theme_minimal()
```

::: {.callout-note}
## Question
An outlier is clearly visible. What is likely the cause? Is it always appropriate
to remove outliers — what factors should you consider before doing so?
:::

```{r remove-outlier}
# tidyverse approach
df_all <- df_all |> filter(height > 7)

# Equivalent base-R approach (shown for comparison)
# df_all <- df_all[df_all$height > 7, ]
```

```{r boxplot-all-clean}
ggplot(df_all) +
  aes(x = "", y = height, fill = birthday) +
  geom_boxplot() +
  scale_fill_hue() +
  labs(x = "Birthday date", y = "Height in cm",
       title = "Boxplot of height by odd/even birthday") +
  theme_minimal()
```

```{r density-all-clean}
ggplot(df_all) +
  aes(x = height, fill = birthday) +
  geom_density(adjust = 1, alpha = 0.5) +
  geom_rug(aes(colour = birthday)) +
  labs(x = "Height in cm", y = "Probability Density",
       title = "Are heights related to birthday dates?",
       fill = "birthday date") +
  theme_minimal()
```

::: {.callout-note}
## Question
After removing the outlier, the two groups overlap considerably. What does
this suggest about the relationship between birthday parity and height?
:::

### 3.5 Comparing with official Japanese height data

```{r jpn-glimpse}
#| eval: false
# The commented-out block below shows how jpn_ht was originally generated —
# shown for transparency, not for running. The data is already available via the URL.

# m_norm  <- round(rnorm(n = 1708, mean = 172,   sd = 5.42), 1)
# f_norm  <- round(rnorm(n = 1559, mean = 158.7, sd = 5.3),  1)
# jpn_ht  <- tibble(id     = sample(1:3267, 3267, replace = FALSE),
#                   sex    = c(rep("male", 1708), rep("female", 1559)),
#                   height = c(m_norm, f_norm)) |>
#   arrange(id)
# write_csv(jpn_ht, "data/jpn_height_survey.csv")
```

```{r jpn-glimpse-live}
glimpse(jpn_ht)
```

```{r jpn-density-combined}
p1 <- ggplot(jpn_ht) +
  aes(x = height) +
  geom_density(adjust = 1, fill = "steelblue") +
  labs(x = "Height in cm", y = "Probability Density",
       title = paste("Density plot of heights in Japan, n =", nrow(jpn_ht))) +
  theme_minimal()
p1
```

::: {.callout-note}
## Question
The plot looks bimodal. Before looking at the next plot, what variable do you
think is driving the two peaks?
:::

```{r jpn-density-by-sex}
p2 <- ggplot(jpn_ht) +
  aes(x = height, fill = sex) +
  geom_density(adjust = 1L, alpha = 0.5) +
  labs(x = "Height in cm", y = "Probability Density") +
  scale_fill_hue() +
  theme(legend.position = "bottom")
p2

gridExtra::grid.arrange(p1, p2)
```

```{r jpn-boxplot}
ggplot(jpn_ht) +
  aes(x = "", y = height, fill = sex) +
  geom_boxplot() +
  scale_fill_hue() +
  labs(y = "Height in cm", title = "Boxplot of heights in Japan") +
  theme(axis.title.x = element_blank())
```

::: {.callout-note}
## Question
With ~3,000 data points, we feel confident that male and female heights
form two distinct populations even without running a formal statistical test.
Why does a larger sample give us more confidence? What test might you use
to confirm this formally?
:::

```{r compare-means}
compare_heights <- tibble(
  class_average  = mean(df_class$height),
  survey_average = mean(df_all$height),
  jpn_average    = mean(jpn_ht$height)
)
compare_heights
```

::: {.callout-note}
## Question
Did the larger all-respondents sample do a better job of matching the
Japanese population average than the class-only sample? What does this
suggest about sample size and representativeness?
:::

---

## Part 4 — Further Monte Carlo examples

### 4.1 Estimating π

The area of a unit circle is π. If we scatter random points uniformly inside a
2×2 square and count how many fall within the unit circle, the ratio gives us an
estimate of π/4.

```{r pi-estimate}
set.seed(314)
n <- 1000000

x <- runif(n, min = -1, max = 1)
y <- runif(n, min = -1, max = 1)
z <- sqrt(x^2 + y^2)   # distance from origin

# Estimate π: 4 × (proportion of points inside the circle)
4 * sum(z <= 1) / length(z)
```

```{r pi-plot}
#| warning: false
InOut <- as.factor(ifelse(z <= 1, "In", "Out"))
plot(x, y, col = InOut, main = "Estimating π with Monte Carlo", pch = ".")
```

::: {.callout-note}
## Questions
- Why does the formula `4 × sum(z <= 1) / n` give an estimate of π?
- How accurate is the estimate? What happens to accuracy as `n` increases?
- Why do we use `runif()` here rather than `rnorm()`?
:::

### 4.2 The Birthday Paradox

::: {.callout-note}
## Background
In a group of just 23 people, the probability that at least two share a birthday
is over 50% — much higher than most people's intuition suggests. R has a built-in
function to compute this mathematically, and we can verify it with simulation.
:::

```{r birthday-builtin}
# R's built-in mathematical solution
pbirthday(n = 23,  coincident = 2)
pbirthday(n = 100, coincident = 2)
```

::: {.callout-note}
## Question
`pbirthday(n = 23)` returns ~0.507. Does that match your intuition?
Why do you think people tend to underestimate this probability?
:::

```{r birthday-simulation}
#| message: false
# Monte Carlo approach: simulate 10,000 trials for each group size
people <- seq(2, 100, 2)   # group sizes: even numbers from 2 to 100
trial  <- 1:10000

# Cross every group size with every trial number
trials <- crossing(trial, people)

# For each row: generate random birthdays and check for duplicates
chance_summary <- trials |>
  mutate(
    birthday = map(people, ~ sample(365, ., replace = TRUE)),
    multiple = map_lgl(birthday, ~ any(duplicated(.)))
  ) |>
  group_by(people) |>
  summarise(chance = mean(multiple))
```

::: {.callout-note}
## Questions
- What does `crossing()` produce? Why do we need every combination of trial and group size?
- What does `map_lgl()` do differently from `map()`?
- Why does `sample(365, ., replace = TRUE)` use `replace = TRUE`?
:::

```{r birthday-plot}
ggplot(chance_summary, aes(people, chance)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Group size",
       y = "Probability two people share a birthday") +
  theme_minimal()
```

```{r birthday-compare}
# Compare simulated probabilities with the mathematical formula
probs <- tibble(
  people     = people,
  math       = map_dbl(people, stats::pbirthday),
  sim        = chance_summary$chance,
  difference = math - sim
)

# Simulated curve
p <- probs |>
  ggplot(aes(people, sim)) +
  geom_line() +
  theme_minimal()
p

# Overlay the mathematical curve in red
p + geom_line(aes(y = math, colour = "mathematical"))
```

```{r birthday-difference}
# How close was the simulation to the maths?
probs |>
  ggplot(aes(people, difference)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Group size", y = "Math − Simulation",
       title = "Difference between mathematical and simulated probabilities") +
  theme_minimal()
```

::: {.callout-note}
## Questions
- How close does the simulation come to the mathematical result?
- Where is the difference largest, and why might that be?
- How would you reduce the difference further?
:::

---

## Further reading

- Descriptive statistics in R, including advanced summary functions:
  <https://www.statsandr.com/blog/descriptive-statistics-in-r/#advanced-descriptive-statistics>
