---
title: "Session 01: Course Introduction: Data Types, R and RStudio"
subtitle: "ISS Data Science — Weekly Notes"
author: "Your Name"
date: "2026-04-14"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    code-download: true
    theme: litera
    self-contained: true
  pdf:
    toc: true
    number-sections: false
  docx:
    toc: true
execute:
  warning: false
  message: false
  error: true        # keeps rendering even if code has errors
---

```{r setup}
#| include: false

# Load packages here — add more as needed
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse)
```

# What I learned this week

*Write a short summary of the main ideas from this session. A few sentences is fine. These are your notes, so Japanese is also completely okay!*

------------------------------------------------------------------------

# Code notes

Use the sections below to paste in code from the lesson and add your own notes. Feel free to rename the headings to match the topics you're covering.

日本語でもオーケー!

## Topic 1

*What does this code do? Write notes / comments above or below the code chunk*

```{r}
# Paste your code here

```

## Topic 2

*What does this code do? Write notes / comments above or below the code chunk*

```{r}
# Paste your code here

```

## Topic 3

*What does this code do? Write notes / comments above or below the code chunk*

```{r}
# Paste your code here

```

------------------------------------------------------------------------

# Things I want to remember

*Anything that surprised you, a useful trick, or something you want to look up later.*

-   

-   

-   

------------------------------------------------------------------------

# Questions I still have

*Anything you didn't understand or want to ask about next session.*

-   

-   

-   

------------------------------------------------------------------------

::: {.callout-tip collapse="true"}
## Rendering tips

**Render to HTML** (recommended): Click the **Render** button in RStudio, or run:

``` r
quarto::quarto_render("my-notes.qmd", output_format = "html")
```

**Switch format**: Change `html` to `pdf` or `docx` in the YAML above, or render all formats at once by removing the format specification and using:

``` r
quarto::quarto_render("my-notes.qmd", output_format = "all")
```

**If a code chunk stops the document from rendering**, add `#| eval: false` at the top of that chunk to skip it:

```` markdown
```{{r}}
#| eval: false
code_that_causes_an_error()
```
````

Or turn off evaluation for the whole document by adding this to the YAML:

``` yaml
execute:
  eval: false
```
:::
