---
title: "Data Project"
author: "WRITE YOUR NAME HERE"
date: today
format:
  html:
    toc: true
    toc-float: true
    embed-resources: true
    df-print: kable
    theme: lumen
    code-download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 16/2, fig.height = 9/2)
```

::: {.callout-note}
### This is a template — adapt it freely

This document is a starting point, not a rigid structure. You are encouraged 
to add sections, rename headings, include extra plots, or reorganise the flow 
to suit your data and research question. What matters is that your final 
document tells a clear, well-supported story.
:::

::: {.callout-important}
### Folder structure

Your project folder is organised like this:

```
data-project/
├── data-project.Rproj
├── data-project.qmd
├── data_raw/          # your original data file — never modify this
├── data_processed/    # optional: save cleaned data here
├── figures/           # save plots here with ggsave()
└── scripts/           # optional: separate R scripts for long tasks
```

Always place your downloaded data in `data_raw/` and read from there in your 
code. Never edit raw data files directly.
:::

```{r libraries}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(ggplot2, dplyr, readr, moderndive)
```

---

# Pre-submission checks

*Complete the Pre sections below and share with your teacher at the deadlines
listed on the project page. You can delete this note from your final submission.*

## Pre_1: Research question

*State your research question clearly. What are you trying to find out, and 
why is it interesting or important?*

**Research question:**

## Pre_2: Data description

*Describe your dataset. Where does it come from? What does it contain? How 
many rows and columns does it have? What are the key variables you plan to use?*

**Data source:**

**Description:**

**Key variables:**

- Outcome variable:
- Numerical predictor(s):
- Categorical predictor(s):

## Pre_3: Load your data

*Load your dataset here. Place your raw data file in `data_raw/` and read it 
from there. Use `glimpse()` to check the structure looks as expected.*

```{r load-data}
# Your code here
# e.g. my_data <- read_csv("data_raw/my-dataset.csv")

```

```{r glimpse, eval=FALSE}
# glimpse(my_data)
```

*Does the data look as expected? Note any immediate issues (wrong variable 
types, unexpected values, missing data).*

---

## Pre_4: Data cleaning

*Clean and prepare your data here. Common tasks include renaming variables, 
filtering out unwanted rows, handling missing values, and creating new variables 
with `mutate()`. Explain what you are doing and why.*

::: {.callout-tip}
If your cleaning steps are very long, consider moving them to a separate script 
in the `scripts/` folder and sourcing it here with `source("scripts/clean.R")`.
:::

```{r clean-data}
# Your code here

```

*Optionally, save your cleaned data to `data_processed/` for future reference:*

```{r save-clean, eval=FALSE}
# write_csv(my_data_clean, "data_processed/my-dataset-clean.csv")
```

## Pre_5: Summary statistics

*Calculate summary statistics for your key variables. Use `summarize()` and 
`group_by()` where appropriate. Comment on anything interesting or unexpected.*

```{r summary-stats}
# Your code here

```

## Pre_6: Exploratory plots

*Create at least two plots that reveal something interesting about your data. 
Each plot should have a title and clear axis labels. Write a sentence or two 
below each plot explaining what it shows.*

```{r plot-1}
# Your code here

```

*What does this plot show?*

```{r plot-2}
# Your code here

```

*What does this plot show?*

```{r save-plots, eval=FALSE}
# Save your plots to the figures/ folder
# ggsave("figures/descriptive-name.png", plot = plot_name, 
#         width = 16/2, height = 9/2)
```

---

# Final write-up

*The sections below are your final submission. Revise and polish your work 
from the Pre sections above. You can delete the Pre sections from your final 
submission if you prefer a cleaner document — or keep them to show your 
process. Either is fine.*

## 1. Introduction

*Write 2–3 paragraphs introducing your project. Cover:*

- *Your research question and why it is interesting*
- *Where your data comes from and what it contains*
- *The key variables you will analyse*

## 2. Exploratory data analysis

*Present your summary statistics and exploratory plots in polished form. 
Each plot should be publication-ready — clear labels, informative title, 
appropriate colours. Write a short interpretation below each one.*

### Summary statistics

```{r final-summary}
# Your code here

```

### Visualisations

```{r final-plot-1}
# Your code here

```

*Interpretation:*

```{r final-plot-2}
# Your code here

```

*Interpretation:*

## 3. Statistical analysis

*Fit at least one regression model to address your research question. Walk 
the reader through your analysis step by step.*

### Visualising the relationship

```{r model-plot}
# Your code here

```

*What does this plot suggest about the relationship between your variables?*

### Fitting the model

```{r model}
# Your code here
# e.g. my_model <- lm(outcome ~ predictor1 + predictor2, data = my_data)

```

### Regression table

```{r regression-table}
# Your code here
# e.g. get_regression_table(my_model)

```

*Interpret the intercept and each slope estimate in plain language. What do 
the numbers actually mean in the context of your data?*

### Residual analysis

```{r residuals}
# Your code here
# e.g. get_regression_points(my_model)

```

*Are there any obvious patterns in the residuals? Any unusually large 
residuals? What does this tell you about the model fit?*

## 4. Discussion

*Interpret your overall findings in relation to your research question. 
Aim for 2–3 paragraphs.*

**What do your results mean?**

**What are the limitations of your analysis?**

**What further questions does your work raise?**

## 5. Citations and references

*List your data sources, R packages used, and any literature cited.*

- **Data:** 
- **Packages:** 
- **References:**

---

*When you are done, render this document to HTML and submit on Moodle. 
Name your file `data-project_yourname.html`. Include this `.qmd` source 
file in your submission.*
